BarnebyLives an R package to create herbarium specimen labels

### 2023-09-08 manuscript/manuscript_APPS.rmd
working on splitting data from a binomial input column

--- 
title: BarnebyLives an R package to create herbarium specimen labels
author:  |
    | Reed Clark Benkendorf$^1$^[Correspondence: rbenkendorf@chicagobotanic.org],  Nyree Zerega$^1$$^,$$^2$, Pat Herendeen$^1$$^,$$^2$, Jeremie B. Fant$^1$$^,$$^2$
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
abstract:  |  
  **Premise:** Depositing quality specimens to herbaria is a time intensive task. Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated to curatorial tasks and specimen deposition. Despite this, the continual generation of herbaria specimens are essential for research in ecology and evolution. In order to faciliate the continued growth of herbaria BarnebyLives was developed as tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks and produce labels.    
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. Census Bureau to provide political jurisdiction information, and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions. It uses inhouse spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclatural innovations. Optionally the package writes driving directions to sites using Google Maps. Finally the package outputs data in a tabular format for review by the user to accept or confirm changes,      
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.     
keywords: |
  Herbarium, collections, natural history museum, geospatial, automation, R, software
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage[width=\textwidth]{caption}
- \usepackage{wrapfig}
---

```{r}
knitr::opts_chunk$set(echo=F, warning = F, message = F)
```

@@

@@

\begin{wrapfigure}{l}{0.4\textwidth}
  \centering
    \includegraphics[width=0.4\textwidth]{../graphics/plots/flowchart.png}
  \caption{Specified benchmarks at all ESDs which included all three metrics and derived imputed value}
\end{wrapfigure}

@@

data <- read.csv('data/test_data.csv', na.strings = "") %>% 
  drop_na(c('longitude', 'latitude', 'date_digital'))

n_families <- data %>% 
  group_by(family) %>% 
  count() %>% 
  nrow() # 74 families

n_spp <- data %>% 
  group_by(binomial) %>% 
  count() %>% 
  nrow() # 616 species

n_infraspecies <- data %>% 
  drop_na(infraspecies) %>% 
  group_by(binomial, infraspecies) %>% 
  count() %>% 
  nrow() # 66 distinct infraspecies

n_sp_authors <- data %>% 
  drop_na(authority) %>% 
  count() # 557 groups of authors

n_infra_sp_auths <- data %>% 
  drop_na(authority1) %>% 
  group_by(binomial, infraspecies) %>% 
  count() %>% 
  nrow() # 22 distinct infra species author groups

@@

time_physical_grabber <- system.time({ # grab sites physical information
  data <- physical_grabber(data, path = p2geo)
}) # not finding slope and aspect

@@

time_geodata_writer <- system.time({ # write out collection as GoogleEarth object
  geodata_writer(data, path = 'someplace', 
               filename = 'Herbarium_Collections_2023',
               filetype = 'kml')
})

@@

# CONCLUSIONS

@@

# ACKNOWLEDGEMENTS
The Bureau of Land Management are graciously acknowledged as providers of funding to R.C.B for the majority of his specimen collection activities. Two anonymous peer reviewers who increased the quality of this manuscript are thanked. Co-collectors who collected specimens used in this study are thanked: Dani Yashinovitz, Dakota Becerra, Hannah Lovell, Hubert Szczygiel, Caitlin Miller.

# DATA AVAILABILITY STATEMENT
The BarnebyLives R package is open source, the development version is available on GitHub (https://github.com/sagesteppe/BarnebyLives), and the stable version is available on CRAN. The package includes three real use-case vignettes (tutorials) on usage. One vignette "setting_up_files" explores setting up a instance for a certain geographic area. Another vignette "running_pipeline" showcases the usage of the package for processing data entered on a spreadsheet. A final vignette "creating_labels" shows the usage of an R, and Bash script launched from RStudio to produce print-ready labels.

@@

# REFERENCES

>>>>

--- 
title: BarnebyLives an R package to create herbarium specimen labels
author:  |
    | Reed Clark Benkendorf$^1$^[Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
abstract:  |  
  **Premise:** Depositing quality specimens to herbaria is a time intensive task. Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated to curatorial tasks and specimen deposition. Despite this, the continual generation of herbaria specimens are essential for research in ecology and evolution. In order to faciliate the continued growth of herbaria BarnebyLives was developed as tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks and produce labels.    
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. Census Bureau to provide political jurisdiction information, and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions. It uses inhouse spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclatural innovations. Optionally the package writes driving directions to sites using Google Maps. Finally the package outputs data in a tabular format for review by the user to accept or confirm changes,      
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.     
keywords: |
  Herbarium, collections, natural history museum, geospatial, automation, R, software
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage[width=\textwidth]{caption}
- \usepackage{wrapfig}
---

```{r echo = F}
knitr::opts_chunk$set(echo=F, warning = F, message = F)
```

@@

\begin{wrapfigure}{l}{0.35\textwidth}
  \centering
    \includegraphics[width=0.35\textwidth]{../graphics/plots/flowchart-trim.png}
  \caption{Specified benchmarks at all ESDs which included all three metrics and derived imputed value}
\end{wrapfigure}

@@

`r lorem::ipsum(paragraphs = 5)`

@@

data <- read.csv('data/test_data.csv', na.strings = "") %>% 
  drop_na(c('Longitude', 'Latitude', 'Date_digital'))

n_families <- data %>% 
  group_by(Family) %>% 
  count() %>% 
  nrow() # 74 families

n_spp <- data %>% 
  group_by(Binomial) %>% 
  count() %>% 
  nrow() # 616 species

n_infraspecies <- data %>% 
  drop_na(Infraspecies) %>% 
  group_by(Binomial, Infraspecies) %>% 
  count() %>% 
  nrow() # 66 distinct infraspecies

n_sp_authors <- data %>% 
  drop_na(Binomial_authority) %>% 
  count() # 557 groups of authors

n_infra_sp_auths <- data %>% 
  drop_na(Infraspecific_authority) %>% 
  group_by(Binomial, Infraspecies) %>% 
  count() %>% 
  nrow() # 22 distinct infra species author groups

@@

time_physical_grabber <- system.time({ # grab sites physical information
  data <- physical_grabber(data, path = p2geo)
})

@@

time_geodata_writer <- system.time({ # write out collection as GoogleEarth object
  geodata_writer(data, path = 'someplace', 
               filename = 'Herbarium_Collections_2023',
               filetype = 'kml')
})
lorem
```

@@

\begin{wrapfigure}{r}{0.45\textwidth}
  \centering
    \includegraphics[width=0.45\textwidth]{../graphics/tables/Screenshot-DataSources.png}
  \caption{Sources of Data required for operations}
\end{wrapfigure}

\newpage
# CONCLUSIONS

`r lorem::ipsum(paragraphs = 3)`

@@

# ACKNOWLEDGEMENTS
The Bureau of Land Management are graciously acknowledged as providers of funding to R.C.B for the majority of his specimen collection activities. Two anonymous peer reviewers who increased the quality of this manuscript are thanked. Co-collectors who collected specimens used in this study are thanked: Dani Yashinovitz, Dakota Becerra, Hannah Lovell, Caitlin Miller & Hubert Szczygiel.

# DATA AVAILABILITY STATEMENT
The BarnebyLives R package is open source, the development version is available on GitHub (https://github.com/sagesteppe/BarnebyLives), and the stable version is available on CRAN. The package includes three real use-case vignettes (tutorials) on usage. One vignette "setting_up_files" explores setting up a instance for a certain geographic area. Another vignette "running_pipeline" showcases the usage of the package for processing data entered on a spreadsheet. A final vignette "creating_labels" shows the usage of an R, and Bash script launched from RStudio to produce print-ready labels. All data used in this mansucript are available at: https://github.com/sagesteppe/Barneby_Lives_dev/manuscript

@@

\newpage 
\small
# REFERENCES
